Cartograms with Rivers

### 2022-11-15 paper/section/appendix.tex
draft: review

\section{Procedure: TestIntersection}
% \begin{noindent}
    \begin{algorithm}[htb!]
        \caption{Procedure to test if a node's translation path, $ \nodeLineNV $ intersects a river.}\label{alg:check river intersection}
        \textbf{Input:} \\
        $ \nodeLineNV \gets $ the node's translation path \\
        $ \river \gets $ a river feature \\

>>>>

\section{Procedure: TranslateRiver}

\begin{algorithm}[tb!]
    \caption{Procedure to translate rivers.}\label{alg:TranslateRiver}
    \textbf{Input:} \\
    $ \nodeList \gets $ a list of $ \node $ representing CCGs \\

    \textbf{Local variables:} \\
    $ count \gets $ the number of river crossing nodes \\
    $ \riverList \gets $ a list of $ \river $ representing river features \\
    $ \node \gets $ a node is an object with the following properties: \\
    \-\hspace{1em} $ \node.x, \node.y, $ or $ \node(x,y) \gets $ the x and y coordinates of $ \node $ \\
    \-\hspace{1em} $ \nodeCross \gets $ the counter for river crosses for $ \node $ \\
    \-\hspace{1em} $ \nodePrevious \gets $ the previous position of $ \node $ \\
    \-\hspace{1em} $ \nodeFNOR \gets $ the translated position of $ \node $ \\

    \begin{algorithmic}[1]
        \Procedure{TranslateRiver}{$ \nodeList $}
        \ForEach {$ \river \in \riverList $}
        \State $ count \gets 0 $
        \State $ \riverVector \gets (0,0) $ \Comment{Hold the sum of vectors $ \nodeVectorCT $}
        \ForEach {$ \node \in \nodeList $}

        \If {$ \node(x,y) \neq \nodeFNOR(x,y) $}
        \If{ \Call{TestIntersection}{$ \nodeFNORLine $, $ \river $} $ = True $}
        \State $ \riverVector \gets \riverVector + \nodeVectorCT$
        \State $ count ++ $
        \EndIf

        \EndIf

        \EndFor

        \State Translate river $ \river $ by the average vector $ \frac{\riverVector}{count} $
        \EndFor

        \EndProcedure
    \end{algorithmic}
\end{algorithm}

\section{Procedure: TestIntersection}
% \begin{noindent}
    \begin{algorithm}[tb!]
        \caption{Procedure to test if a node's translation path, $ \nodeLineNV $ intersects a river.}\label{alg:check river intersection}
        \textbf{Input:} \\
        $ \nodeLineNV \gets $ the node's translation path \\
        $ \river \gets $ a river feature \\

### 2022-11-15 paper/section/conclusion.tex
draft: review

In this paper, we present a novel approach to generate cartograms with rivers. We first propose a new algorithm to generate cartograms with rivers, and then we present a prototype to support the exploration of cartograms with rivers. We evaluate our approach using a user study, and the results partially support our hypotheses: the introduction of rivers improves the legibility but doesn't reduce the time needed to locate a region. We also discuss the limitations of our work and the future directions.

>>>>

In this paper, we present a novel approach to generate non-contiguous cartograms with rivers. We first propose a new algorithm to generate cartograms with rivers, and then we present a prototype to support the exploration of cartograms with rivers. We evaluate our approach using a user study, and the results  support our hypotheses: the introduction of rivers improves the legibility and recognizability of cartograms, although a deeper analysis is necessary to reach more confident conclusions. We also discuss the limitations of our work and future research directions.

### 2022-11-15 paper/section/data.tex
draft: review

{
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography Portalx \cite{opengeographyportalxOpen} with EPSG:4326 (WGS84 - World Geodetic System) as the Coordinate Reference System (CRS).}
    \label{fig:ccg}
\end{figure}
}

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The first step is to obtain both geospatial boundaries and EHR data. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for cartograms.

Geospatial boundaries, or shapefiles, were obtained from the sources described here.

@@

\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}

@@

For all datasets, a spreadsheet including the following data is provided:

>>>>

 {
  \begin{figure}[tb!]
      \centering
      \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
      \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography Portalx \cite{opengeographyportalxOpen} with EPSG:4326 (WGS84 - World Geodetic System) as the Coordinate Reference System (CRS).}
      \label{fig:ccg}
  \end{figure}
 }

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The first step is to obtain both geospatial boundaries and EHR data. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for cartograms. Geospatial boundaries, or shapefiles, were obtained from the sources described here.

@@

\begin{lstlisting}[float=tp,caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}

@@

For all datasets, a spreadsheet including the following is provided:

### 2022-11-15 paper/section/evaluation.tex
draft: review

\newcommand{\pCount}{15 }
\section{User-Centered Evaluation}

@@

\textbf{H2:} The introduction of rivers can reduce the time needed to locate a CCG in a cartogram.

We test these hypotheses using location based tasks. Given a standard choropleth on the left side of the screen and a cartogram on the right side, we ask user study participants to locate the given region in the choropleth in the cartogram.

@@

\subsubsection{Independant Variables}

The primary independent variable is the presence of rivers in the cartograms, which directly impacts the final layout of the cartogram. We set this parameter to either render rivers or not.

@@

\subsubsection{Dependant Variables}

\bobgraph{Accuracy:} Given a CCG location in the choropleth on the left, we ask the participant to locate the corresponding node in the cartogram. The is the primary dependent variable measured by the number of correct CCGs chosen by the participants.

@@

\subsubsection{Control Variables}
\bobgraph{Choice of color map:} We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green, to depict the data in our cartograms.

@@

\subsection{User Study}
In this section we describe the user study participants, datasets, and the experiment procedure.

@@

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/task.png}
        \caption{A sample task for participants. The left shows the choropleth map, and the right shows the corresponding cartogram. Both images show the three longest rivers in England, and the color is mapped to CCG population. The target CGG is blinking on the choropleth (shown in black), participants are asked to identify this CCG on the cartogram.}
        \label{fig:task}
    \end{figure}
}

@@

The user study is designed to be conducted online, and participants are asked to complete the tasks using their own computer. The user study includes four parts:

\textbf{P1:} The participants are asked to listen to instructions and training provided as both text and videos. The instructions are designed to help participants understand the concepts used in the tasks. Instructional videos are available at \url{https://www.youtu.be/playlist?list=PLL7sHvxLtD75fMtrUQrAdddjt3wfFkcWz}.

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design. A demonstration of three sample tasks is also included in the instructional video.

\textbf{P3:} The participants are asked to complete 16 location tasks. The response and reaction time are recorded. These 16 CCGs were carefully selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

@@

These are provided in the appendix, and available at \url{https://osf.io/q39w7}.

@@

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/task-rt.png}
        \caption{The scatter plot shows the response time (in seconds) and answers for each task. Correct answers are shown in green circle, and incorrect answers are shown in red triangle. Solid glyphs represent the task without rivers, and hollow glyphs represent the task with rivers.}
        \label{fig:task-all}
    \end{figure}
}

\bobgraph{Accuracy:} \Cref{fig:task-acc} shows the accuracy measured by the number of correct CCGs chosen by the participants. The chart shows that the introduction of rivers has improved the accuracy for two CCGs. The accuracy for the two groups differs significantly according to the two-sample unequal variance $t$-test performed, $t$(176.43) = 2.92, $p$ = 0.0039. This supports our hypothesis H1 that the introduction of rivers improves the legibility of cartogram.

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/accuracy.png}
        \caption{The bar chart shows the accuracy of locating CCGs.}
        \label{fig:task-acc}
    \end{figure}
}

\bobgraph{Response Time:} \Cref{fig:task-rt} shows the average response time measured for four CCGs. The chart shows that the introduction of rivers has slightly reduced the time needed to locate CCGs. The two groups show no significant differences according to the two-sample unequal variance $t$-test performed, $t$(168.03) = 1.29, $p$ = 0.20.  The statistical analysis does not support our hypothesis H2, that the introduction of rivers does not reduce the response time significantly.

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/rt.png}
        \caption{The bar chart shows the average response time for locating CCGs.}
        \label{fig:task-rt}
    \end{figure}
}

>>>>

\newcommand{\pCount}{24 }
\section{User-Centered Evaluation}

@@

\textbf{H1.1:} The introduction of rivers can increase target location accuracy.

\textbf{H1.2:} The introduction of rivers can reduce the time needed to locate a CCG in a cartogram.

We test these hypotheses using location based tasks. Given a standard choropleth on the left side of the screen and a cartogram on the right side, we ask user study participants to locate the given region in the choropleth in the cartogram. See \Cref{fig:task}.

@@

\bobgraph{Independant Variables: }The primary independent variable is the presence of rivers in the cartograms, which directly impacts the final layout of the cartogram. We set this parameter to either render rivers or not.

@@

\bobgraph{Dependant Variables: }\bobgraph{Accuracy:} Given a CCG location in the choropleth on the left, we ask the participant to locate the corresponding node in the cartogram. The is the primary dependent variable measured by the number of correct CCGs chosen by the participants.

@@

\bobgraph{Control Variables: }\bobgraph{Choice of color map:} We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green, to depict the data in our cartograms.

@@

\subsection{User Study Design}
In this section we describe the user study participants, datasets, and the experiment procedure.

@@

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/task.png}
    \caption{A sample task for participants. The left shows the choropleth map, and the right shows the corresponding cartogram. Both images show the three longest rivers in England, and the color is mapped to CCG population. The target CGG is blinking on the choropleth (shown in black), participants are asked to identify this CCG on the cartogram.}
    \label{fig:task}
\end{figure}
}

@@

The user study is designed to be conducted online, and participants are asked to complete the tasks using their own hardware. The user study includes four parts:

\textbf{P1:} The participants are asked to listen to instructions and training provided as both text and videos. The instructions are designed to help participants understand the concepts used in the tasks. Instructional videos are available at \URL{https://www.youtu.be/playlist?list=PLL7sHvxLtD75fMtrUQrAdddjt3wfFkcWz}.

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design. These tasks also serve as a quality test to see if the users are taking the study seriously. A demonstration of three sample tasks is also included in the instructional video.

\textbf{P3:} The participants are asked to complete 16 location tasks. The response and reaction time are recorded. These 16 CCGs were selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

@@

The entire study is available at \URL{https://osf.io/q39w7}.

@@

\bobgraph{Accuracy:} \Cref{fig:task-acc} shows the accuracy measured by the number of correct CCGs chosen by the participants. The chart shows that the introduction of rivers has improved the accuracy for two CCGs. The accuracy for the two groups differs significantly according to the two-sample unequal variance $t$-test performed, $t$(176.43) = 2.92, $p$ = 0.0039. This supports our hypothesis H1.1 that the introduction of rivers increases target location accuracy.

    {
        \begin{figure}[tb!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/accuracy.png}
            \caption{The bar chart shows the accuracy of locating CCGs with and without rivers.}
            \label{fig:task-acc}
        \end{figure}
    }

\bobgraph{Response Time:} \Cref{fig:task-rt} shows the average response time measured for four CCGs. The chart shows that the introduction of rivers has slightly reduced the time needed to locate CCGs. The two groups show no significant differences according to the two-sample unequal variance $t$-test performed, $t$(168.03) = 1.29, $p$ = 0.20. The statistical analysis does not support our hypothesis H1.2, that the introduction of rivers reduces the response time significantly.

    {
        \begin{figure}[tb!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/rt.png}
            \caption{The bar chart shows the average response time for locating CCGs with and without rivers.}
            \label{fig:task-rt}
        \end{figure}
    }

\bobgraph{Feedback:} \Cref{fig:likert} shows the results of the following Likert Scale questions:
\begin{enumerate}[label=(\Alph*),align=left,leftmargin=*,labelindent=1em]
    \item 87.50\% of participants agree that including rivers is useful.
    \item 83.33\% of participants agree that rivers increase the legibility of a cartogram.
    \item 95.83\% of participants agree that including rivers makes cartograms easier to understand.
    \item 62.50\% of participants agree that including rivers makes CCGs easier to locate.
    \item 87.50\% of participants agree that including rivers adds value to the standard cartogram.
\end{enumerate}

    {
        \begin{figure}[b!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/likert.png}
            \caption{The stacked bar chart shows the responses of Likert Scale questions.}
            \label{fig:likert}
        \end{figure}
    }

### 2022-11-15 paper/section/introduction.tex
draft: review

{
    \begin{figure*}[ht!]
        \centering
        \includegraphics[width=0.8\textwidth,keepaspectratio]{figure/UI.png}
        \caption{A screenshot of the software. User options are provided to adjust the size, color mapping, visibility of nodes and rivers. Other options include the ability to control the overlap removal behavior: rivers can be static, dynamic, or constrained [why too complicated? this is what we are doing in the software].}
        \label{fig:overview}
    \end{figure*}
}

@@

Cartograms are representations of hybrid geographical and abstract data based on a value-by-area mapping combining statistical and geographical information \cite{dent2009Cartography}. Various styles of cartograms have been proposed and implemented, covering applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and public health decision-making \cite{gao2020Visualising, sack2021Visualizing}.

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \Cref{table:accuracy}). For this project, we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. Standard Demers cartograms are composed of uniform shape nodes. As such, this can reduce their legibility. We implement a new cartographic layout algorithm that includes rivers into the layout of the nodes representing geographical regions. Introducing rivers into the cartogram improves their legibility. To reduce geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real-world case study using an Electronic Health Records (EHR) dataset to evaluate the result. We present a user study demonstrating its effectiveness.

>>>>

@@

Cartograms are representations of hybrid geographical and abstract data based on a value-by-area mapping combining statistical and geographical information \cite{dent2009Cartography}. Various styles of cartograms have been proposed and implemented, covering applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and decision-making for public health \cite{gao2020Visualising, sack2021Visualizing}.

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \Cref{table:accuracy}). For this project, we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. Standard Demers cartograms are composed of uniform shape nodes. As such, this can reduce their legibility. We implement a new cartographic layout algorithm that includes rivers into the placement of the nodes representing geographical regions. Introducing rivers into the cartogram improves their legibility. To reduce geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real-world case study using Electronic Health Records (EHR) data to evaluate the result. We present a user study demonstrating its effectiveness.

### 2022-11-15 paper/section/limitation.tex
draft: review

In this section we discuss some limitations of our work and the future directions. The first limitation is the color map we use to depict the data in our cartograms. We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green. However, we believe that the choice of color map can have a significant impact on the legibility of cartogram. In the user study, we carefully avoid extreme values where the location or color of the CCG makes it easier to locate the target, see \Cref{fig:extreme}. We plan to explore the impact of different color maps on the legibility of cartogram in future work.

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/limitations/extreme.png}
        \caption{Due to color and relative location, the CCGs in the black circle are extremely easy to locate.}
        \label{fig:extreme}
    \end{figure}
}

>>>>

In this section we discuss some limitations of our work and future research directions. The first limitation is the color map we use to depict the data in our cartograms. We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green. However, we believe that the choice of color map can have a significant impact on the legibility of cartograms. In the user study, we carefully avoid extreme values where the location or color of the CCG makes it easier to locate the target, see \Cref{fig:extreme}. We plan to explore the impact of different color maps on the legibility of cartogram in future work.

{
    \begin{figure}[tb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/limitations/extreme.png}
        \caption{Due to color and relative location, we believe the CCGs in the black circle are easier to locate.}
        \label{fig:extreme}
    \end{figure}
}
